---
title: "Cyclistic Casual Users vs Members"
author: "Felipe"
date: "2023-02-17"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

### Data collection and combination into a single file

Using the read.csv() function, the last 12 months of the Cyclistic bike
sharing system's usage were imported into R.
After verifying consistency in the column headers for all 12 documents, they were joined into a single dataframe named "all_trips".

```{r loading the documents, include=FALSE}
library(ggplot2)
library(lubridate)
library(dplyr)
setwd("C:/Users/Felipe/Desktop/Docs/DA/Case Study") 
feb_2022 <- read.csv("Cyclistic’s historical trip data/Original documents (CSV files)/202202-divvy-tripdata.csv")
mar_2022 <- read.csv("Cyclistic’s historical trip data/Original documents (CSV files)/202203-divvy-tripdata.csv")
apr_2022 <- read.csv("Cyclistic’s historical trip data/Original documents (CSV files)/202204-divvy-tripdata.csv")
may_2022 <- read.csv("Cyclistic’s historical trip data/Original documents (CSV files)/202205-divvy-tripdata.csv")
jun_2022 <- read.csv("Cyclistic’s historical trip data/Original documents (CSV files)/202206-divvy-tripdata.csv")
jul_2022 <- read.csv("Cyclistic’s historical trip data/Original documents (CSV files)/202207-divvy-tripdata.csv")
aug_2022 <- read.csv("Cyclistic’s historical trip data/Original documents (CSV files)/202208-divvy-tripdata.csv")
sep_2022 <- read.csv("Cyclistic’s historical trip data/Original documents (CSV files)/202209-divvy-tripdata.csv")
oct_2022 <- read.csv("Cyclistic’s historical trip data/Original documents (CSV files)/202210-divvy-tripdata.csv")
nov_2022 <- read.csv("Cyclistic’s historical trip data/Original documents (CSV files)/202211-divvy-tripdata.csv")
dec_2022 <- read.csv("Cyclistic’s historical trip data/Original documents (CSV files)/202212-divvy-tripdata.csv")
jan_2023 <- read.csv("Cyclistic’s historical trip data/Original documents (CSV files)/202301-divvy-tripdata.csv")

all_trips <- bind_rows(feb_2022, mar_2022, apr_2022, may_2022, jun_2022, jul_2022, aug_2022, sep_2022, oct_2022, nov_2022, dec_2022, jan_2023)

```

```{r dataframe combination}
glimpse(all_trips)
```

### Data cleanup and preparation for the analysis

The data can only be aggregated at the ride-level, which is too granular. Additional columns for month, day and year were added.


```{r}
all_trips$date <- as.Date(all_trips$started_at) 
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")                    
all_trips$year <- format(as.Date(all_trips$date),"%y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```

In addition, a new column "ride_length" was created by subtracting the difference between the end-time and the start-time. The time is presented in seconds.

```{r}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at) 
```

```{r include=FALSE}
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)

```

The dataframe includes a few hundred entries when bikes were taken out of docks for maintenance. Moreover, some ride lengths appear negative.

Since information was being removed, an updated version of the dataframe called "all_trips_v2" was created.

```{r}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]

```


### Descriptive Analysis

#### Ride length
```{r}
summary(all_trips_v2$ride_length)
```

#### Comparing members and casual users

Average ride time for members and casual users
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
```

Median ride length for members and casual users
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
```

Maximum ride length for members and casual users
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
```

Minimum ride length for members and casual users
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
```

Average ride time for members and users each day of the week
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

Ridership data filtered and sorted by type and day of the week
```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%                  #groups by user-type and weekday
  summarise(number_of_rides = n()				                #calculates the number of rides and average duration 
  ,average_duration = mean(ride_length)) %>% 		        # calculates the average duration
  arrange(member_casual, weekday)		
```

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, month) %>%                  #groups by user-type and weekday
  summarise(number_of_rides = n()				                #calculates the number of rides and average duration 
  ,average_duration = mean(ride_length)) %>% 		        # calculates the average duration
  arrange(member_casual, month)		
```

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, month) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, month)  %>% 
  ggplot(aes(x = month, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") + ggtitle("Average Daily Rides") + xlab("Month") + ylab("Average rides") + labs(fill="User type")
```



Number of rides by user type
```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") + ggtitle("Average Monthly Rides") + xlab("Day of the week") + ylab("Amount of rides") + labs(fill="User type")

```

Trips average duration
```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration , fill = member_casual)) +
  geom_col(position = "dodge") + ggtitle("Trip Average Duration") + xlab("Day of the week") + ylab("Time (s)") + labs(fill="User type")

```

